---
title: NCAH Quantative Data Analysis from 2019 to 2024
execute: 
  echo: false
---

```{r}
#| label: library
#| output: false
library(tidyverse)
library(janitor)
library(knitr)
library(showtext)
library(ggrepel)
library(svglite)
font_add_google(name = "Roboto", family = "Roboto")
showtext_auto()
```
Below is our data analysis conducted for the North Carolina Alliance for Health on the participation of schools in the Commmunity Eligible Provision (CEP) in North Carolina.

## Data Read in

```{r}
#| label: read-2022-24
#| output: false
cep_2024 <- read_csv("data/2024_CEP_USDA_Annual Notification_NCDPI.csv") |> 
  mutate(year = 2024) |> 
  rename("identified_student_percentage_isp" = "Identified Student Percentage (ISP)")

cep_2022 <- read_csv("data/2022_USDA CEP Annual Notification report_NCDPI.csv") |> 
  rename("Currently Participating in CEP in One or Some Schools (Not District-wide)" = 
           "Currently Participating in CEP in One or Some Schools\n(Not District-wide)") |> 
  mutate(year = 2022) |> 
  rename(
    "identified_student_percentage_isp" = 
      "Identified Student Percentage\r\n(ISP)"
  )

cep_2023 <- read_csv("data/NCDPI CEP Notification - April 2023.csv") |> 
  mutate(year = 2023) |> 
    rename(
    "identified_student_percentage_isp" = 
      "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2024)
```

We collected our data from the NCDPI Annual Notifications of School Eligibility Report from 2019 through 2024. We read in all the data and binded the rows together.

In addition, we mutated a new column that indicated the form of participation of each ditrict, alongside cleaning the identified_student_percentage_isp variable to turn it into a double variable

```{r}
#| label: read-2021-data
#| output: false
cep_2021 <- read_csv("data/2021 NCDPI CEP.csv") |> 
  mutate(year = 2021) |> 
  rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2021)
```

```{r}
#| label: read-2020-data
#| output: false
cep_2020 <- read_csv("data/2020 NCDPI CEP.csv") |> 
  mutate(year = 2020) |> 
  rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2020)
```

```{r}
#| label: read-2019-data
#| output: false
cep_2019 <- read_csv("data/2019 CEP NCDPI.csv") |> 
  mutate(year = 2019) |> 
  rename("identified_student_percentage_isp" = "Identified Student Percentage (ISP)") |> 
  rename("LEA name" = "LEA Name")
glimpse(cep_2019)
```

```{r}
#| label: bind-all-rows
#| output: false
cep_all_raw <- cep_2019 |> 
  bind_rows(cep_2020, cep_2021, cep_2022, cep_2023, cep_2024) |> 
  select(-...15, -Comments) |> 
  clean_names()
```

```{r}
#| label: cep-clean
#| output: false
cep_all <- cep_all_raw |>   
  mutate(
    participation_type = case_when(
      currently_participating_in_cep_district_wide == "A" ~ "District Wide",
      currently_participating_in_cep_in_one_or_some_schools_not_district_wide == "S" ~ "Some Schools",
      .default = "Not Participating"
    ),
    participation_type = fct_relevel(participation_type, "District Wide", "Some Schools", "Not Participating")
  ) |> 
  mutate(
    identified_student_percentage_isp = str_remove(identified_student_percentage_isp, "[%]"),
    identified_student_percentage_isp = as.double(identified_student_percentage_isp)
  )

cep_all |> 
  write_csv("data/cep-all.csv")

cep_yearly <- cep_all |> 
  select(lea_name, year, participation_type, identified_student_percentage_isp) |> 
  drop_na()

cep_yearly |> 
  pivot_wider(names_from = year, values_from = participation_type) |> 
  write_csv("data/cep-yearly.csv")
  
```

## District Participation in CEP

```{r}
#| label: district-cep-graph
#| fig-asp: .7
participation_count <- cep_all |> 
  count(year, participation_type) |> 
  drop_na()

participation_count |> 
  ggplot(aes(x = year, y = n, color = participation_type)) +
  geom_line(linewidth = 1.25) +
  geom_vline(xintercept = 2023, alpha = .35,) +
  labs(
    x = "Year",
    y = "Number of Districts",
    color = "Participation Type"
  ) +
  scale_color_manual(values = c("#1e3c7c", "#0580cd", "#e97132")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    text = element_text(family = "Roboto"),
    plot.background = element_rect(fill = "#f7f7f7", size = 0)
  )
ggsave("graphs/district_wide_participation.svg")
```

```{r}
#| label: tbl-data-count
#| tbl-cap: "District Participation in CEP by Year"
participation_count |> 
  pivot_wider(names_from = year, values_from = n) |> 
  rename(`Participation Type` = participation_type) |> 
  kable()
```

## District Wide ISP and Participation in CEP Program

```{r}
#| label: fig-isp-density-viz
#| fig-asp: .6
cep_all |> 
  filter(year == 2024) |> 
  ggplot(
    aes(x = identified_student_percentage_isp, 
        fill = participation_type,
        colour = participation_type)
  ) +
  geom_density(alpha = .1) +
  geom_vline(xintercept = 62.5, alpha = .45) +
  labs(
    x = "Identified Student Percentage (ISP)",
    y = "Density",
    color = "Participation Type",
    fill = "Participation Type"
  ) +
  scale_fill_manual(values = c("#1e3c7c", "#0580cd", "#e97132")) +
  scale_color_manual(values = c("#1e3c7c", "#0580cd", "#e97132")) +
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(family = "Roboto"),
        plot.background = element_rect(fill = "#f7f7f7", size = 0))
ggsave("graphs/isp_density.svg")
```

```{r}
#| label: fig-isp-participation
#| fig-asp: .6
#| fig-cap: "ISP Participation over 62.5% leads to extremly high full district participation in the Community Eligable Provision"

cep_all |> 
  mutate(
    fed_funding = if_else(identified_student_percentage_isp >= 62.5,
            "Full Funding",
            "Partial Funding")) |> 
  count(fed_funding, participation_type) |> 
  drop_na() |> 
  ggplot(aes(
    x = fed_funding, y = n, fill = participation_type, 
  )) +
  geom_col(position = "fill") +
  labs(
    x = "Federal Funding",
    y = "Proportion of Districts",
    fill = "Participation Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  ) +
  scale_fill_discrete(c("#0D3A4F", "#FA824C", "#CE272A"))
```

```{r}
#| label: fig-student-counts-in-cep-schools

cep_all
not_cep_2023 <- cep_all |> 
  filter(year == 2023 & participation_type == "Not Participating") |> 
  select(lea_id, lea_name, identified_student_percentage_isp, year, participation_type)
not_cep_2023

yes_cep_2024 <- cep_all |> 
  filter(year == 2024 & (participation_type == "Some Schools" | 
                           participation_type == "District Wide")) |> 
  select(lea_id, lea_name, identified_student_percentage_isp, year, participation_type)
yes_cep_2024

cep_all |> 
  filter(year == 2023 & participation_type != "Not Participating") |> 
  summarize(mean(identified_student_percentage_isp))

bind_rows(not_cep_2023, yes_cep_2024) |> 
  group_by(lea_id) |> 
  filter(n() == 2) |>
  ungroup() |> 
  summarize(mean(identified_student_percentage_isp))
```

