---
title: NCAH Quantative Data Analysis from 2019 to 2024
execute: 
  echo: false
---

```{r}
#| label: library
#| output: false
library(tidyverse)
library(janitor)
```

## Data Read in

```{r}
#| label: read-2022-24
#| output: false
cep_2024 <- read_csv("data/2024_CEP_USDA_Annual Notification_NCDPI.csv") |> 
  mutate(year = 2024) |> 
  rename("identified_student_percentage_isp" = "Identified Student Percentage (ISP)")

cep_2022 <- read_csv("data/2022_USDA CEP Annual Notification report_NCDPI.csv") |> 
  rename("Currently Participating in CEP in One or Some Schools (Not District-wide)" = "Currently Participating in CEP in One or Some Schools\n(Not District-wide)") |> 
  mutate(year = 2022) |> 
  rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )

cep_2023 <- read_csv("data/NCDPI CEP Notification - April 2023.csv") |> 
  mutate(year = 2023) |> 
    rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2024)
```

We then work in adding our new data
```{r}
#| label: read-2021-data
#| output: false
cep_2021 <- read_csv("data/2021 NCDPI CEP.csv") |> 
  mutate(year = 2021) |> 
  rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2021)
```

```{r}
#| label: read-2020-data
#| output: false
cep_2020 <- read_csv("data/2020 NCDPI CEP.csv") |> 
  mutate(year = 2020) |> 
  rename(
    "identified_student_percentage_isp" = "Identified Student Percentage\r\n(ISP)"
  )
glimpse(cep_2020)
```

```{r}
#| label: read-2019-data
#| output: false
cep_2019 <- read_csv("data/2019 CEP NCDPI.csv") |> 
  mutate(year = 2019) |> 
  rename("identified_student_percentage_isp" = "Identified Student Percentage (ISP)") |> 
  rename("LEA name" = "LEA Name")
glimpse(cep_2019)
```

## Bind rows

With all of our data read in, we binded the six datasets together.
```{r}
#| label: bind-all-rows
#| output: false
cep_all_raw <- cep_2019 |> 
  bind_rows(cep_2020, cep_2021, cep_2022, cep_2023, cep_2024) |> 
  select(-...15, -Comments) |> 
  clean_names()
cep_all_raw
```
We mutated a new column that indicated the form of participation of each ditrict, alongside cleaning the identified_student_percentage_isp variable to turn it into a double variable
```{r}
#| label: cep-clean
#| output: false
cep_all <- cep_all_raw |>   
  mutate(
    participation_type = case_when(
      currently_participating_in_cep_district_wide == "A" ~ "District Wide",
      currently_participating_in_cep_in_one_or_some_schools_not_district_wide == "S" ~ "Some Schools",
      .default = "Not Participating"
    ),
    participation_type = fct_relevel(participation_type, "District Wide", "Some Schools", "Not Participating")
  ) |> 
  mutate(
    identified_student_percentage_isp = str_remove(identified_student_percentage_isp, "[%]"),
    identified_student_percentage_isp = as.double(identified_student_percentage_isp)
  )

cep_all
```

## District Participation in CEP

```{r}
#| label: district-cep-graph
participation_count <- cep_all |> 
  count(year, participation_type)

participation_count |> 
  ggplot(aes(x = year, y = n, color = participation_type)) +
  geom_line() +
  geom_vline(xintercept = 2023, alpha = .35) +
  labs(
    x = "Year",
    y = "Number of Districts",
    color = "Participation Type",
    title = "Changes in CEP Participation by District",
    subtitle = "New ISP thresholds in 2023 increased CEP participation from 62 to 111"
  ) +
  theme_minimal()
```

```{r}
#| label: data-count
participation_count |> 
  pivot_wider(names_from = year, values_from = n)
```

## District Wide ISP and Participation in CEP Program

```{r}
#| label: isp-participation
#| fig-asp: .6
cep_all |> 
  filter(year == 2024) |> 
  ggplot(
    aes(x = identified_student_percentage_isp, 
        fill = participation_type,
        colour = participation_type)
  ) +
  geom_density(alpha = .2) +
  geom_vline(xintercept = 62.5, alpha = .45) +
  labs(
    x = "Identified Student Percentage (ISP)",
    y = "Density",
    color = "Participation Type",
    fill = "Participation Type",
    title = "Relationship between Identified Student Percentage and Participation"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

cep_all |> 
  mutate(
    fed_funding = if_else(identified_student_percentage_isp >= 62.5,
            "full funding",
            "partial funding")) |> 
  count(fed_funding, participation_type) |> 
  drop_na() |> 
  ggplot(aes(
    x = fed_funding, y = n, fill = participation_type, 
  )) +
  geom_col(position = "fill") +
  labs(
    x = "Federal Funding",
    y = "Proportion of Districts",
    fill = "Participation Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )
```



